Typing a linear \picalc

### 2020-03-24 paper/paper.tex
Minor typesetting tweaks
Add some comments on capabilities

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\setlist{parsep=0pt,listparindent=\parindent}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

\newcommand{\subst}[3]{#1[#2/#3]}
\newcommand{\lamcalc}{$\lambda$-calculus}
\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\types}[4]{#1 \propto #2 \vdash #3 \boxtimes #4}
\newcommand{\contains}[5]{#1 \propto #2 \ni #3 \propto #4 \boxtimes #5}
\newcommand{\suc}{{\scriptscriptstyle 1+}}
\newcommand{\op}[3]{#1 := #2 \cdot #3}

@@

\subsection{Multiplicities}

A type system with both linear and shared resources has multiplicities $0$, $1$ and $\omega$.

\begin{figure}[h]
\begin{equation}
\begin{split}
  \constr{0\cdot}          &: C \\
  \constr{+\cdot}          &: C \\
  \constr{-\cdot}          &: C \\
  \constr{\op{\_}{\_}{\_}} &: C \to C \to C \to Set \\
  \constr{1\cdot}          &: C \\
  \constr{\cdot-join}      &: \forall \{x y z\} \to \op{x}{y}{+\cdot} \to \op{x}{z}{-\cdot} \to \exists w . (\op{x}{w}{1\cdot}) \\
  \constr{\cdot-compute}   &: \forall y z \to Dec (\exists x . (\op{x}{y}{z})) \\
  \constr{\cdot-compute^l} &: \forall x z \to Dec (\exists y . (\op{x}{y}{z})) \\
  \constr{\cdot-unique}    &: \forall \{x x' y z\} \to \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  \constr{\cdot-unique^l}  &: \forall \{x y y' z\} \to \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  \constr{\cdot-id^l}      &: \forall x \to \op{x}{0\cdot}{x} \\
  \constr{\cdot-comm}      &: \forall \{x y z\} \to \op{x}{y}{z} \to \op{x}{z}{y} \\
  \constr{\cdot-assoc}     &: \forall \{x y z u v\} \to \op{x}{y}{z} \to \op{y}{u}{v} \to \exists w . (\op{x}{u}{w} \times \op{w}{v}{z})
\end{split}
\end{equation}
\caption{Partial commutative monoid}
\end{figure}
\todo{alignment, get rid of $1\cdot$, $\cdot-join$, $\cdot-compute^l$}

\begin{figure}[h]
\begin{equation}
\begin{split}
  \constr{\Idx}          &: Set \\
  \constr{\exists \Idx}  &: Idx \\
  \constr{\Carrier}      &: \Idx \to Set \\
  \constr{\Quantifiers}  &: \forall i : \Idx \to \Quantifier_{\Carrier_i}
\end{split}
\end{equation}
\caption{Indexed set of partial commutative monoids}
\end{figure}

@@

Shared. Gradual. Affine. Linear.

@@

\subsection{Typing with Leftovers}

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\setlist{parsep=0pt,listparindent=\parindent}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage{mathtools}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

\newcommand{\subst}[3]{#1[#2/#3]}
\newcommand{\lamcalc}{$\lambda$-calculus}
\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\types}[4]{#1 \propto #2 \vdash #3 \boxtimes #4}
\newcommand{\contains}[5]{#1 \propto #2 \ni #3 \propto #4 \boxtimes #5}
\newcommand{\suc}{{\scriptscriptstyle 1+}}
\newcommand{\op}[3]{#1 \coloneqq #2 \cdot #3}

@@

A type system is in the business of further constraining a language, filtering out terms that (although well-formed) are meaningless.
In the simply typed \lamcalc{}, a term has a meaning iff it has a type.
This typing judgment is a proof of the form $\Gamma \vdash t : T$, where the term $t$ has type $T$ under the typing context $\Gamma$ -- the typing context associates variables (possibly occurring within $t$) to types.
In the simply typed \picalc{} however, it is the channels that are typed, and a process has a meaning iff it \emph{uses} its channels according to their types.
The typing judgment is therefore of the form $\Gamma \vdash p$, where the process $p$ (which has no type of its own) uses the channels in $\Gamma$ according to their types.

A resource-aware type system uses both types and usage counting to constraint a language.
It uses typing contexts to keep track of both types and usage counts.
In the case of a linear type system, resources must be used exactly once, and usages transition from available to consumed as soon as they are used.
In the linear \picalc{} however, every channel must be used exactly once to receive and exactly once to send.
In \S \ref{multiplicities}, we generalize this usage counting algebra so that the same type system can be used for a combination of linear, shared, gradual and affine types.

Syntax directed
Continuations.
\ref{leftover-typing}

\subsection{Multiplicities}\label{multiplicities}

In its usual representation, the linear \picalc{} distinguishes between \emph{capabilities} -- $\ell_{\o}$ for cannot send or receive, $\ell_i$ for input, $\ell_o$ for output, and $\ell_\#$ for both input and output -- and multiplicities -- $0$ or $1$.
This introduces unnecessary semantic duplication: the semantic meaning of a multiplicity of $0$ on a capability $c$ is independent of $c$; a multiplicity of 1 $\ell_\#$ is equivalent to a multiplicity of 1 $\ell_i$ and 1 $\ell_o$.
In this paper, we merge multiplicities and capabilities into a single set and provide an algebra that generalizes the rules of the linear \picalc{} into rules that are suitable for shared, linear, affine and gradual typing.

The algebra $\Quantifier_C$ on the set of capabilities $C$ presented in Figure \ref{capabilities} is based on a \emph{partial commutative monoid} with a few special elements.
The operation of the partial monoid is defined as a relation $\op{x}{y}{z}$ on a carrier $C$.
This relation must be commutative, associative, and absorb the neutral element $\ell_{\o}$.
To recover the computability of the relation, $\constr{\cdot-compute}$ requires that given two elements $y$ and $z$, it is decidable whether a third element $x$ exists such that $\op{x}{y}{z}$.
To recover the functionality, $\constr{\cdot-unique}$ requires that given any two elements $y$ and $z$, if a third one exists, it is uniquely defined.
A consequence of using leftover typing (\S \ref{leftover-typing}) is that this algebra needs to be injective, and thus $\constr{\cdot-unique^l}$ requires for an operand to be uniquely defined as well.
In addition to these rules, we define the special elements $\ell_i$ and $\ell_o$, to be subtracted every time an input or output occurs, respectively.
Finally, we require for $\ell_i$ and $\ell_o$ to be compatible, and refer to the result as $\ell_\#$.

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\ell_{\o}                &:{} &                      &     & C \\
  &\ell_i                   &:{} &                      &     & C \\
  &\ell_o                   &:{} &                      &     & C \\
  &\op{\_}{\_}{\_}          &:{} &                      &     & C \to C \to C \to Set \\
  &\constr{\cdot-join}      &:{} &                      &     & \exists \ell_\# . \; \op{\ell_\#}{\ell_i}{\ell_o} \\
  &\constr{\cdot-compute}   &:{} &\forall y z           & \to & Dec \; (\exists x . \; (\op{x}{y}{z})) \\
  &\constr{\cdot-unique}    &:{} &\forall \{x x' y z\}  & \to & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  &\constr{\cdot-unique^l}  &:{} &\forall \{x y y' z\}  & \to & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  &\constr{\cdot-id^l}      &:{} &\forall x             & \to & \op{x}{\ell_{\o}}{x} \\
  &\constr{\cdot-comm}      &:{} &\forall \{x y z\}     & \to & \op{x}{y}{z} \to \op{x}{z}{y} \\
  &\constr{\cdot-assoc}     &:{} &\forall \{x y z u v\} & \to & \op{x}{y}{z} \to \op{y}{u}{v} \to \exists w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
\end{aligned}
\end{equation}
\caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid with special elements $\ell_i$, $\ell_o$ and $\ell_\#$.}
\label{capabilities}
\end{figure}

To allow for several capability algebras to coexist on the same type system, in Figure \ref{indexed-capabilities} we index the algebra carriers by an index $\Idx$.
Such an index is a type with one element for each algebra of our type system.
To be able to prove subject congruence, we require for at least one such algebra to exist.
Finally, the field $\constr{\Quantifiers}$ requires for each index $i$ to have an algebra defined on the carrier set $\constr{\Carrier_i}$.

\begin{figure}[h]
\begin{equation}
\begin{split}
  \constr{\Idx}          &: Set \\
  \constr{\exists \Idx}  &: Idx \\
  \constr{\Carrier}      &: \Idx \to Set \\
  \constr{\Quantifiers}  &: \forall i : \Idx \to \Quantifier_{\Carrier_i}
\end{split}
\end{equation}
\caption{Indexed set of partial commutative monoids}
\label{indexed-capabilities}
\end{figure}

@@

\paragraph{Shared}

The carrier $C$ is implemented as a type $\constr{One}$ with a single trivial constructor $\omega$.
All of $\ell_{\o}$, $\ell_i$, $\ell_o$ and $\ell_\#$ are instantiated to $\omega$.
The relation $\op{\omega}{\omega}{\omega}$ is inhabited.
All other laws are trivially satisfied.

\paragraph{Linear}

The carrier $C$ is implemented as a type $\constr{Four}$ with the trivial elements $\ell_{\o}$, $\ell_i$, $\ell_o$ and $\ell_\#$.
The partial monoid is defined with the element $\ell_{\o}$ being neutral on both sides, and the element $\ell_\#$ splitting into $\ell_i$ and $\ell_o$ or $\ell_o$ and $\ell_i$ -- and is uninhabited in every other case.
All other rules follow trivially.

\paragraph{Affine}

The algebra of an affine system is exactly the one of a linear one, the only difference is the output context must satisfy a trivial predicate.

\paragraph{Gradual}

The carrier $C$ is implemented as a tuple $(\N , \N)$ of natural numbers.
The element $\ell_{\o}$ corresponds to the tuple $(0 , 0)$, $\ell_i$ to $(1 , 0)$, $\ell_o$ to $(0 , 1)$, and $\ell_\#$ to $(1 , 1)$.
The partial monoid is defined exactly when $\op{(x_l + y_l , x_r + y_r)}{(x_l , x_r)}{(y_l , y_r)}$.
All other rules follow trivially from the algebraic rules for the addition of naturals.

@@

\subsection{Typing with Leftovers}\label{leftover typing}

@@

\todo{comment on scope restriction rule and capabilities}
